Instant Continuation Marks--Just Add $\lambda$

### 2012-09-28 paper.tex
Alpha.

For example, the ubiquitous JavaScript is an ideal candidate for the addition of continuation marks. However, as the \emph{lingua franca} of the web, it has numerous mature implementations which have been heavily optimized. To add continuation marks to JavaScript amounts to modifying each implementation upstream, to say nothing of amending the JavaScript standard. (Clements et al. successfully added continuation marks in Mozilla's Rhino compiler \cite{clements2008implementing}, but it remains a proof-of-concept.)

@@

Continuation marks respect tail-call optimization (TCO). A tail call in a language with TCO does not enlarge the continuation. Instead, the previous portion of the execution context is subsumed by the new. Conceptually, the continuation does not grow to accommodate new marks; if one is made, the previous is overwritten. Thus, with a tail-recursive definition of the factorial function as in
\begin{schemedisplay}
(define (fac n acc)
  (if (zero? n)
      (begin
        (print (ccm))
        acc)
      (wcm n (fac (- n 1) (* n acc)))))
\end{schemedisplay}
evaluation of \scheme|(fac 2)| proceeds as
\begin{schemedisplay}
(fac 2 1)
(if (zero? 2) ... (wcm 2 (fac (- 2 1) (* 2 1))))
(wcm 2 (fac (- 2 1) (* 2 1)))
(wcm 2 (fac 1 (* 2 1)))
(wcm 2 (fac 1 2))
(wcm 2 (if (zero? 1) ... (wcm 1 (fac (- 1 1) (* 1 2)))))
(wcm 2 (wcm 1 (fac (- 1 1) (* 1 2))))
(wcm 1 (fac (- 1 1) (* 1 2)))
(wcm 1 (fac 0 (* 1 2)))
(wcm 1 (fac 0 2))
(wcm 1 (if (zero? 0) (begin (print (ccm)) 2) ...))
(wcm 1 (begin (print (ccm)) 2))
(wcm 1 (begin (print '(1)) 2))
'(1)
(wcm 1 (begin 2))
(wcm 1 2)
2
\end{schemedisplay}
and the tail-call behavior is clear.

@@

All together, the transformation adds three arguments to every term $e$: the continuation $k$ representative of the current context $E$; a flag $f$ encoding whether evaluation is occurring with the body of a \scheme'wcm' expression; and the transformed value of $\chi$ for the current context $\C{\chi(E)}$.

@@

The definition over \scheme'wcm' expressions is the only one to make explicit use of the value of the flag. Following the evaluation of the mark expression, rule \ref{collapse} is applied, if necessary. The continuity of $\chi$ allows us to construct $\chi(E'[\wcm{v_0}{e_1}])$ from $\chi(E)$ where

\begin{equation}
E'=\begin{cases}
E' &\text{if $E=E'[\wcm{v'}{\hole}]$ for some $E'$ and $v'$}\\
E &\text{otherwise}
\end{cases}
\end{equation}

@@

\setspecialsymbol{lambda}{$\lambda$}
\setspecialsymbol{->}{$\to$}
\setspecialsymbol{-->}{$\rightarrow$}
\setspecialsymbol{betav}{$\beta$v}
\setconstant{error}
\setkeyword{chi chi_p}

@@

@@

>>>>

For example, the ubiquitous JavaScript is an ideal candidate for the addition of continuation marks. However, as the lingua franca of the web, it has numerous mature implementations which have been heavily optimized. To add continuation marks to JavaScript amounts to modifying each implementation upstream, to say nothing of amending the JavaScript standard. (Clements et al. successfully added continuation marks in Mozilla's Rhino compiler \cite{clements2008implementing}, but it remains a proof-of-concept.)

@@

Continuation marks respect tail-call optimization. A tail call in a tail-call optimized language does not enlarge the continuation. Instead, the previous portion of the execution context is subsumed by the new. Conceptually, the continuation does not grow to accommodate new marks; if one is made, the previous is overwritten. Thus, with a tail-recursive definition of the factorial function as in
\begin{schemedisplay}
(define (fac n acc)
  (if (zero? n)
      (begin
        (print (ccm))
        acc)
      (wcm n (fac (- n 1) (* n acc)))))
\end{schemedisplay}
evaluation of \scheme|(fac 2)| proceeds as
\begin{schemedisplay}
(fac 2 1)
(if (zero? 2) ... (wcm 2 (fac (- 2 1) (* 2 1))))
(wcm 2 (fac (- 2 1) (* 2 1)))
(wcm 2 (fac 1 (* 2 1)))
(wcm 2 (fac 1 2))
(wcm 2 (if (zero? 1) ... (wcm 1 (fac (- 1 1) (* 1 2)))))
(wcm 2 (wcm 1 (fac (- 1 1) (* 1 2))))
(wcm 1 (fac (- 1 1) (* 1 2)))
(wcm 1 (fac 0 (* 1 2)))
(wcm 1 (fac 0 2))
(wcm 1 (if (zero? 0) (begin (print (ccm)) 2) ...))
(wcm 1 (begin (print (ccm)) 2))
(wcm 1 (begin (print '(1)) 2))
'(1)
(wcm 1 (begin 2))
(wcm 1 2)
2
\end{schemedisplay}
and the tail-call behavior is clear.

@@

All together, the transformation adds three arguments to every term $e$: the continuation $k$ representative of the current context $E$; a flag $f$ encoding whether evaluation is occurring with the body of a \scheme'wcm' expression; and the transformed value $m$ of $\chi$ for the current context $\C{\chi(E)}$.

@@

The definition over \scheme'wcm' expressions is the only one to make explicit use of the value of the flag. Following the evaluation of the mark expression, rule \ref{tail-enforce} is applied, if necessary. The continuity of $\chi$ allows us to construct $\chi(E''[\wcm{v_0}{e_1}])$ from $\chi(E)$ where

\begin{equation}
E''=\begin{cases}
E' &\text{if $E=E'[\wcm{v'}{\hole}]$ for some $E'$ and $v'$}\\
E &\text{otherwise}
\end{cases}
\end{equation}

@@

\setspecialsymbol{lambda}{$\lambda$}
\setspecialsymbol{->}{$\to$}
\setspecialsymbol{-->}{$\rightarrow$}
\setspecialsymbol{betav}{$\beta$v}
\setconstant{error}
\setkeyword{chi chi_p number apply-reduction-relation*}

@@

\setkeyword{meaning-preserved? alpha-eq? transform eval init redex-check}

@@

\begin{definition}
\[
\mathrm{eval}_{cm}(p)=\begin{cases}
v     &\text{if $p\cmrrs v$}\\
\perp &\text{if $p\cmrrs\cdots$}
\end{cases}
\]
\end{definition}

\begin{definition}
\[
\mathrm{eval}_{v}(p)=\begin{cases}
v     &\text{if $p\lvrrs v$}\\
\perp &\text{if $p\lvrrs\cdots$}
\end{cases}
\]
\end{definition}

\newtheorem{maintheorem}{Correctness of $\mathcal{C}$}
\begin{maintheorem}
For all programs $p\in\lambda_{cm}$, $\C{\mathrm{eval}_{cm}(p)}=\mathrm{eval}_{v}(\C{p})$.
\end{maintheorem}

\section{Back to JavaScript}

With a correct transformation of continuation marks defined over the \lc, it requires little to add continuation marks to an eager, higher-order language such as JavaScript.

We add two keywords to manipulate continuation marks: the familiar \texttt{wcm} and \texttt{ccm}. We chose block syntax for \texttt{wcm} to convey that it is a special form, like a conditional, instead of a function-like interface. A lone \texttt{ccm} evaluates to the current continuation marks.

Factorial in this JavaScript extension can be implemented like the following:
\begin{verbatim}
var fac = function( n ) {
    if( n == 0 ) {
        console.log( ccm );
        return 1;
    }
    else {
        return wcm { n }
                   { n * fac( n - 1 ) };
    }
}
\end{verbatim}

A transformation of this function desugars these constructs into valid JavaScript.
\begin{verbatim}
var fac = function( f, m ) {
    return function( n ) {
        if( n == 0 ) {
            console.log( (function( f, m ) { return m; })( false, m ) );
            return 1;
        }
        else {
            return (function( f, m ) {
                return (function( v ) {
                    return n * fac( false, [ v ].concat( f ? m.slice( 1 ) : m ) )( n - 1 );
                })( n );
            })( f, m );
        }
    }
}
\end{verbatim}
